---
title: "Exporing Data in the The Internet Movie Database"
output: html_document
---
## Introduction
In this project, I explore the `movies` data set from the `ggplot2movies` package. The data is from the internet movie database, [imdb.com](http://imdb.com). It's one of the largest movie databases today and is a collection of movie data supplied by studios and fans. There are 58,788 rows and 24 variables in this data set.
```{r, echo=FALSE}
library(ggplot2movies)
data(movies)
```

## Exploratory Data Analysis
I develop some questions that come to mind and that I think are interesting to know. I also keep in mind what kind of visualizations I can create based upon these questions and answers.

What is the range of years of production of the movies of this data set (i.e. what is the year of production of the oldest movie and of the most recent movie in this data set)?
```{r}
min(movies$year) #year of the oldest movie
max(movies$year) #year of the most recent movie
```
The year of production of the oldest movie is 1893 and the year of production of the most recent movie in this data set is 2005.

What proportion of movies have their budget included in this data set, and what proportion doesn’t?
What are top 5 most expensive movies in this data set?
```{r}
budgetNAs <- sum(is.na(movies$budget)) #number of NAs in the budget column
totalMovies <- nrow(movies) #total number of rows (movies)
budgetNAs/totalMovies #proportion of movies w/ NO budget included
1-budgetNAs/totalMovies #proportion of movies w/ budget included

#top 5 most expensive movies
movies[head(order(movies$budget,decreasing = T),5),c("title","budget")]
```
About 91.1% of movies have their budget included in this data set and the rest, 8.9%, do not. The top 5 most expensive movies in this data set are "Spider-Man 2", "Titanic", "Troy", "Terminator 3: Rise of the Machines", and "Waterworld".

What are top 5 longest movies?
```{r}
#top 5 longest movies
movies[head(order(movies$length,decreasing = T),5),c("title","length")]
```
The top 5 longest movies are The Cure for Insomnia, The Longest Most Meaningless Movie in the World, Four Stars, Resan, and Out 1.

Of all short movies, which one is the shortest (in minutes)? Which one is the longest? How long are the
shortest and the longest short movies?
```{r}
moviesShort <- movies[movies$Short==1,] #filtered data with short films only

#shortest short film
moviesShort[head(order(moviesShort$length),1),c("title","length")]

#number of short films w/ length of 1 minute
sum(moviesShort$length==1)

#longest short film
moviesShort[head(order(moviesShort$length,decreasing = T),1),c("title","length")]
```
The shortest short film is 17 Seconds to Sophie with a length of 1 minute. However, I noticed that there are a total of 165 short films with a length of 1 minute. 17 Seconds to Sophie just happens to be at the top of the list because it's in alphabetical order. The longest short film is 10 jaar leuven kort with a length of 240 minutes (which could arguably not be a short film).

How many movies of each genre are there in this data set?
```{r}
#create new aggregated data frame by summing binary genre columns
moviesGenre <- data.frame("Genre" = c("Action","Animation","Comedy","Drama",
                                      "Documentary","Romance","Short"),
                          "Count" = c(sum(movies$Action),sum(movies$Animation),
                                      sum(movies$Comedy),sum(movies$Drama),
                                      sum(movies$Documentary),sum(movies$Romance),
                                      sum(movies$Short)))
moviesGenre
#create barplot
xx <- barplot(height = moviesGenre$Count,
              names.arg = moviesGenre$Genre,
              las = 2,
              col = topo.colors(length(moviesGenre$Genre)),
              ylim = c(0,25000),
              main = "Number of Movies by Genre")
#add data labels on top of each bar
text(x = xx, y = moviesGenre$Count, label = moviesGenre$Count, pos = 3)
```

What is the average rating of all movies within each genre?
```{r}
#create new aggregated data frame by averaging rating column if value in each
#genre column is equal to 1
moviesRating <- data.frame("Genre" = c("Action","Animation","Comedy","Drama",
                                      "Documentary","Romance","Short"),
                          "Avg Rating" = c(mean(movies[movies$Action==1,]$rating),
                                      mean(movies[movies$Animation==1,]$rating),
                                      mean(movies[movies$Comedy==1,]$rating),
                                      mean(movies[movies$Drama==1,]$rating),
                                      mean(movies[movies$Documentary==1,]$rating),
                                      mean(movies[movies$Romance==1,]$rating),
                                      mean(movies[movies$Short==1,]$rating)))
#round 2 decimal places
moviesRating$Avg.Rating <- round(moviesRating$Avg.Rating,2)
moviesRating
#create barplot
yy <- barplot(height = moviesRating$Avg.Rating,
              names.arg = moviesRating$Genre,
              las = 2,
              col = topo.colors(length(moviesRating$Genre)),
              ylim = c(0,8),
              main = "Average Rating of Movies by Genre")
#add data labels on top of each bar
text(x = yy, y = moviesRating$Avg.Rating, label = moviesRating$Avg.Rating, pos = 3)
```

What is the average rating of all movies within each genre that were produced in the years 2000-2005?
```{r}
#create new aggregated data frame by averaging rating column if value in each
#genre column is equal to 1
moviesRating2000_2005 <- data.frame("Genre" = c("Action","Animation","Comedy","Drama",
                                      "Documentary","Romance","Short"),
                          "Avg Rating" = c(mean(movies[movies$Action==1
                                                       & movies$year>=2000
                                                       & movies$year<="2005",]$rating),
                                           mean(movies[movies$Animation==1
                                                       & movies$year>=2000
                                                       & movies$year<="2005",]$rating),
                                           mean(movies[movies$Comedy==1
                                                       & movies$year>=2000
                                                       & movies$year<="2005",]$rating),
                                           mean(movies[movies$Drama==1
                                                       & movies$year>=2000
                                                       & movies$year<="2005",]$rating),
                                           mean(movies[movies$Documentary==1
                                                       & movies$year>=2000
                                                       & movies$year<="2005",]$rating),
                                           mean(movies[movies$Romance==1
                                                       & movies$year>=2000
                                                       & movies$year<="2005",]$rating),
                                           mean(movies[movies$Short==1
                                                       & movies$year>=2000
                                                       & movies$year<="2005",]$rating)))
#round 2 decimal places
moviesRating2000_2005$Avg.Rating <- round(moviesRating2000_2005$Avg.Rating,2)
moviesRating2000_2005
#create barplot
zz <- barplot(height = moviesRating2000_2005$Avg.Rating,
              names.arg = moviesRating2000_2005$Genre,
              las = 2,
              col = topo.colors(length(moviesRating2000_2005$Genre)),
              ylim = c(0,8),
              main = "Average Rating of Movies Produced from 2000-2005 by Genre")
#add data labels on top of each bar
text(x = zz, y = moviesRating2000_2005$Avg.Rating,
     label = moviesRating2000_2005$Avg.Rating, pos = 3)
```

For this part, I wanted to combine multiple plots on one graph. For each genre (not including short films) plot movies from 1990-2005 and plot a function of the number of movies in this data set of corresponding genre produced by year, for years from 1990 until the last year recorded.

```{r}
#create empty dataframe with first column as years
movies1990_2005 <- data.frame("Year" = 1990:2005)
#add genres as column names
movies1990_2005[c("Action","Animation","Comedy","Drama",
                  "Documentary","Romance")] <- NA
#for each genre column, loop through each year and add up the number of movies
for (yr in 1990:2005) {
  count <- sum(movies[movies$year==yr,]$Action)
  movies1990_2005[yr-1989,"Action"] <- count
}
for (yr in 1990:2005) {
  count <- sum(movies[movies$year==yr,]$Animation)
  movies1990_2005[yr-1989,"Animation"] <- count
}
for (yr in 1990:2005) {
  count <- sum(movies[movies$year==yr,]$Comedy)
  movies1990_2005[yr-1989,"Comedy"] <- count
}
for (yr in 1990:2005) {
  count <- sum(movies[movies$year==yr,]$Drama)
  movies1990_2005[yr-1989,"Drama"] <- count
}
for (yr in 1990:2005) {
  count <- sum(movies[movies$year==yr,]$Documentary)
  movies1990_2005[yr-1989,"Documentary"] <- count
}
for (yr in 1990:2005) {
  count <- sum(movies[movies$year==yr,]$Romance)
  movies1990_2005[yr-1989,"Romance"] <- count
}
movies1990_2005
#plot
plot(1, type="n", xlab="Year", ylab="Number of Movies",
     xlim=c(1990,2005), ylim=c(0, 1000), las = 2,
     main = "Number of Movies by Genre from 1990 to 2005")
axis(1, at=seq(1990, 2005, 1), las = 2)
lines(movies1990_2005$Year,movies1990_2005$Action,
      col = topo.colors(length(moviesRating2000_2005$Genre))[1],
      lwd = 2)
lines(movies1990_2005$Year,movies1990_2005$Animation,
      col = topo.colors(length(moviesRating2000_2005$Genre))[2],
      lwd = 2)
lines(movies1990_2005$Year,movies1990_2005$Comedy,
      col = topo.colors(length(moviesRating2000_2005$Genre))[3],
      lwd = 2)
lines(movies1990_2005$Year,movies1990_2005$Drama,
      col = topo.colors(length(moviesRating2000_2005$Genre))[4],
      lwd = 2)
lines(movies1990_2005$Year,movies1990_2005$Documentary,
      col = topo.colors(length(moviesRating2000_2005$Genre))[5],
      lwd = 2)
lines(movies1990_2005$Year,movies1990_2005$Romance,
      col = topo.colors(length(moviesRating2000_2005$Genre))[6],
      lwd = 2)
#legend
legend("topleft",legend=colnames(movies1990_2005)[2:7],
       fill=topo.colors(length(moviesRating2000_2005$Genre)),bty='n')
```

For this last section, I utilize ggplot to visulize the correlation between variables.
  
For movies that are not short films produced from 2000-2005, analyze the correlation between budget and rating by plotting budget on the y-axis and rating on the x-axis.
```{r setup, warning=FALSE}
#create dataframe omitting rows where budget is NA
moviesBudget2000_2005 <- movies[!is.na(movies$budget) &
                                  movies$budget!=0 &
                                  movies$year>=2000 &
                                  movies$year<=2005 &
                                  movies$Short==0 &
                                  movies$mpaa!="",]
#plot
library(ggplot2)
g <-ggplot(data = moviesBudget2000_2005,aes(x = budget, y = rating,
                                            color = mpaa)) + geom_point()
g + geom_smooth(method="lm", level = 0.99)
```

For all mpaa ratings, as budget increases, rating slightly increases as well. You can see this because all three of the regression lines are slightly positive. There are more rated R movies with lower budgets because there are way more blue points on the left side of the graph than any other color. The most expensive movie in this data set was a PG-13 movie with a rating that is quite high at about 7.9. This is shown as the green point at the far top right of the plot.